---
title: "Revisiting the "
author: "Ashton Baker"
output:
  html_document:
    toc: yes
    toc_depth: 4
---

\newcommand\prob[1]{\mathbb{P}\left[{#1}\right]}
\newcommand\expect[1]{\mathbb{E}\left[{#1}\right]}
\newcommand\var[1]{\mathrm{Var}\left[{#1}\right]}
\newcommand\dist[2]{\mathrm{#1}\left(#2\right)}
\newcommand\dlta[1]{{\Delta}{#1}}
\newcommand\lik{\mathcal{L}}
\newcommand\loglik{\ell}

[Licensed under the Creative Commons Attribution-NonCommercial license](http://creativecommons.org/licenses/by-nc/4.0/).
Please share and remix noncommercially, mentioning its origin.  


```{r knitr-opts,include=FALSE,purl=FALSE}
library(knitr)
prefix <- "mif"
opts_chunk$set(
  progress=TRUE,
  prompt=FALSE,
  tidy=FALSE,
  highlight=TRUE,
  strip.white=TRUE,
  warning=FALSE,
  message=FALSE,
  error=FALSE,
  echo=TRUE,
  cache=TRUE,
  cache.extra=rand_seed,
  results='markup',
  fig.show='asis',
  size='small',
  fig.path=paste0("figure/",prefix,"-"),
  cache.path=paste0("cache/",prefix,"-"),
  fig.align='center',
  fig.height=4,fig.width=6.83,
  dpi=100,
  dev='png',
  dev.args=list(bg='transparent')
)
options(keep.source=TRUE,encoding="UTF-8")
```

----------------------------

## Introduction
In this tutorial, we will cover the implementation of models used in Dennis et al. (2001) using the **pomp** software package.


## apsdif

```{r prelims, include = TRUE, purl=FALSE}
library(ggplot2)
library(plyr)
library(reshape2)
library(pomp)
library(magrittr)
library(reshape2)
library(foreach)
stopifnot(packageVersion("pomp")>="1.8")
set.seed(557976883)
```

We begin by importing the data for the experiment. We will initially load only the data for rep 4 into the **pomp** object. Later, we will describe a function that incorporates all of the data to estimate parameters.

```{r lpa-data, include = TRUE}
read.csv("./data/data.csv") %>%subset(weeks <= 40, select=c(weeks,rep,L_obs,P_obs,A_obs)) -> dat
head(dat)
```


```{r lpa-model, include = TRUE}
LPA_model <- pomp(data = subset(dat, rep==4),
              times = "weeks", t0 = 0);
```

## LPA model
Individuals are categorized as either larvae, pupae, or adults. At time $t$, the number of each is given by $L_t$, $P_t$ and $A_t$, respectively. The time unit is 2 weeks, which is the average amount of time spent in the larval stage in this experiment.

Eggs are produced at a rate proportional to the number of adults. Eggs are cannibalized by larvae at a rate $c_{el}$, and by adults at a rate $c_{ea}$. So, the probability of an egg surviving 1 time unit in the presence of $L$ larvae is $\exp(-c_{el}L)$. Likewise, the probability of an egg surving 1 time unit in the presence of $A$ adults is $\exp(-c_{ea}A)$.

Larvae have a mortality rate of $\mu_l$ over the course of 2 weeks. If they survive, we assume they become pupae. Pupae are cannibalized by adults at a rate of $-c_{pa}$. So the probability that a pupa will survive 1 time unit in the presence of A adults is $\exp(-c_{pa} A)$. If they survive, we assume they become adults. - Adults have a mortality rate of $\mu_A$ over the course of 1 time unit.

Thus, the full LPA model is

$$\begin{aligned}
L_t &= b A_{t-1} \exp{(-c_{el} L_{t-1} - c_{ea} A_{t-1})} \\
P_t &= L_{t-1} (1 - \mu_1) \\
A_t &= P_{t-1} \exp{(-c_{pa} A_{t-1})} + A_{t-1}(1 - \mu_a)
\end{aligned}$$

We can use this as the deterministic skeleton of a **pomp** model:

```{r lpa-skeleton, include = TRUE}
skel <- Csnippet(' DL = b * A * exp(-cel * L - cea * A);
                   DP = L * (1 - ul);
                   DA = P * exp(-cpa * A) + A * (1 - ua);');
LPA_model <- pomp(LPA_model,
    skeleton = map(skel, delta.t=2),
    statenames = c('L', 'P', 'A'),
    paramnames = c('b', 'cel', 'cea', 'cpa', 'ua', 'ul'));
```

Each experimental repetition was initialized by placing 250 larvae, 5 pupae, and 100 adults in a jar. So we simply initialize our process with the following `Csnippet`:

```{r lpa-init, include=TRUE}
init <- Csnippet(' L = 250;
                   P = 5;
                   A = 100;');

LPA_model <- pomp(LPA_model, initializer = init, statenames = c('L', 'P', 'A'))
```

We can take the maximum likelihood (ML) estimates for the parameters from the paper, and then we have enough to compute trajectories of the deterministic skeleton:

```{r lpa-ml, include=TRUE}
ml_estimate <- c(b = 10.67, cel = 0.01647, cea = 0.01313, cpa = 0.01313, ua = 0.007629, ul = 0.1955)

plot(L~time, data=trajectory(LPA_model, params = ml_estimate, as.data.frame = TRUE), type = 'l')
plot(P~time, data=trajectory(LPA_model, params = ml_estimate, as.data.frame = TRUE), type = 'l')
plot(A~time, data=trajectory(LPA_model, params = ml_estimate, as.data.frame = TRUE), type = 'l')
```

## Stochastic Demographic LPA Model

To incorporate stochasticity to our model, we need to specify a function for $f_{X_n | X_{n - 1}}$. Dennis et al begin with the LPA model specified above, and modify it as follows.

* $L_t$ is Poisson distributed with mean $b a_{t-1} \exp{(-c_{el} l_{t-1} - c_{ea} a_{t-1})}$.
* Each larva is given a probability $(1 - \mu_l)$ of surviving a 2-week period, at which point we assume it becomes a pupa. Therefore, $P_t$ is binomial distributed with $n = l_{t-1}$ and $p = (1 - \mu_l)$.
* Over a 2-week period, we assume that each pupa either survives to become an adult, or dies. Since pupae are cannibalized by adults, the survival probaility is $\exp(-c_{pa}a_{t-1})$. The survivors are called "recruits", and given by $R_t$.
* Over a 2-week period, we assume that adult beetles survive with probability $(1 - \mu_a)$. Thus, the number of survivors $S_t$ is binomially distributed.

$$\begin{aligned}
L_t &= \text{Poisson}(ba_{t-1} \exp[-c_{el}l_{t-1} - c_{ea}a_{t-1}])\\
P_t &= \text{binomial}(l_{t - 1}, (1 - \mu_l)) \\
R_t &= \text{binomial}(p_{t-1}, \exp[-c_{pa}a_{t-1}]) \\
S_t &= \text{binomial}(a_{t-1}, [1 - \mu_a]) \\
A_t &= R_t + S_t
\end{aligned}$$

We can write a Csnippet to simulate this process as follows:
```{r, include=TRUE}
rproc <- discrete.time.sim( 
         step.fun=Csnippet(' L = rpois(b * A * exp(-cel * L - cea * A));
                             P = rbinom(L, (1 - mu_L));
                             double R = rbinom(P, exp(-cpa * A));
                             double S = rbinom(A, (1 - mu_A));
                             A = R + S;'),
         delta.t=2);
```

We assume the measurement error is negligable, and therefore make no attempt to model it. As a result, our `rmeasure` function simply assigns the value of the $L$, $P$, and $A$ states to the respective observation values.

```{r, include=TRUE}
rmeas <- Csnippet("
                    L_obs = L;
                    P_obs = P;
                    A_obs = A;");
```

## NLAR Modifications
$$\begin{aligned}
\sqrt{L_t} &= \sqrt{bA_{t-1} \exp(-c_{el} L_{t-1} - c_{ea} A_{t-1})} + E_{1t}\\
\sqrt{P_t} &= \sqrt{L_{t-1}(1 - \mu_l)} + E_{2t}\\
\sqrt{A_t} &= \sqrt{P_{t-1} exp(-c_{pa}A_{t-1}) + A_{t-1}(1 - \mu_a)} + E_{3t}\\
\end{aligned}$$

## References